Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2022-05-09 2022/eamt/eamt-latex/eamt22.tex
try to fix style

\author{Minh-Quang Pham\\
  Universit\'e Paris-Saclay, \\
  CNRS, LISN,  91405 Orsay, France \\
  {\tt minhquang.pham@limsi.fr}  \\ \\
  \textbf{Josep Crego}\\
  SYSTRAN ,\\
  5 rue Feydeau, 75002 Paris, France\\
  {\tt josep.crego@systrangroup.com} \And
  Fran\c cois Yvon\\
  Universit\'e Paris-Saclay, \\
  CNRS, LISN,  91405 Orsay, France\\
  {\tt francois.yvon@limsi.fr}}

@@

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for the task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key aspect of training is then to mitigate the detrimental effects of a mismatch between these distributions. Single-source and multi-source\footnote{In this paper, multi-source DA means having multiple domains to adapt from; this setting differs from multi-source translation, where several \emph{source languages} are considered.} domain adaptation (DA) is a well-studied instance of this setting (see \newcite{Chu2017comparison}; \newcite{Saunders21asurvey} for a review), and so is multi-domain (MD) learning \cite{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Aharoni19massively},\fyDone{Fix ref}
% \revision{cite also \cite{Shen21source}} -> I think this is not so appropriate here
where the diversity of training data not only corresponds to variations in the topic, genre, or register but also in language.

@@

Several recent proposals
% \cite{Vanderwees17dynamic,Graves17automated,platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Wang20balancing,Wang20learning-multi}
explore ways to instead consider \emph{dynamic} data selection and sampling strategies: \newcite{Vanderwees17dynamic}; \newcite{Zhang19curriculum} construct a static curriculum, while \newcite{Wang20learning-multi}; \newcite{Wang20balancing} build curricula that automatically adapt to the training data. In this paper, we contribute to this line of research in several ways. 
\begin{itemize}
	\item First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short), a variant of Differentiable Data Selection (DDS) of \newcite{Wang20balancing}, initially applied to multilingual NMT, that simultaneously accounts for the domain adaptation and the multidomain adaptation problems.
	\item We show that MDAC achieves performance that compare to fine-tuning strategies for DA (\textsection~\ref{ssec:da}) and outperform some static data sampling strategies for multidomain settings (\ref{ssec:mda}).
	\item We show that our variant MDAC mitigates some failures of DDS in multidomain training.
	\item We illustrate the generality of differentiable data selection frameworks (both MDAC and DDS) on less common situations such as DA using unsupervised clustering (\textsection~\ref{ssec:clda}); DA using out-of-domain training data and small in-domain validation data (\textsection~\ref{ssec:uda}); and two-domain adaptation where the test distribution only mixes two of the training domain (\textsection~\ref{ssec:bida}).
\end{itemize}

@@

These situations have been amply documented from a theoretical perspective (eg.\ \newcite{Mansour09multiple}; \newcite{Mansour09domainadaptation}; \newcite{Hoffman18algorithms}). A general recommendation in the DA setting is to adjust the sampling distribution used to optimize the system so as to compensate for the mismatch between $\mathcal{D}^s(x)$ and $\mathcal{D}^t(x)$. This can be approximated by reweighting instances, or more conveniently domains, which are selected during training with a probability $\lambda^{l}(d)$, with $\lambda^{l}(d) \neq \lambda^{s}(d)$.

@@

Multidomain NMT (MDMT) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDMT combine one or several ingredients: (a) the specialization of data representations \cite{Kobus17domaincontrol} or of sub-networks \cite{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \cite{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \newcite{Jiang19multidomain}. Unsupervised MDMT is studied in \newcite{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive/dynamic data selection take inspiration from \newcite{Bengio09curriculum}, where the notion of curriculum learning is introduced. CL relies on the notion of the ``easiness'' of a sample to schedule the presentation of training data so that the easiest examples are presented first and the hardest last. Various ways to automate CL using the framework of multi-armed bandits are explored in \cite{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for domain adaptive / multidomain / multilingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}
For instance, \newcite{Zhang19curriculum} study supervised DA and propose a curriculum approach which progressively augments the training data: early stages only use in-data, while less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is opposite to the policy of \newcite{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively increases the focus on in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\newcite{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the current data selection strategy.\fyDone{Alert: what do we do during warm-up ?} This technique is applied to multilingual NMT in \cite{Kumar21learning}. \newcite{Zhou20uncertainty} propose another curriculum-based approach which relies on \emph{instance uncertainty} as a measure of their difficulty and presents data samples starting with the easiest (more predictable) first. Another contribution of this paper is an alternative stopping criterium. Closest to our problems, \newcite{Wang20learning-multi} adapt CL for multidomain NMT, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the dev set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform a variety of adaptation tasks for machine translation, ranging from conventional supervised domain adaptation to multidomain NMT and unseen domain adaptation. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during the course of training, can be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.

@@

% \section*{Acknowledgments}
%The work presented in this paper was partially sup-ported by the European Commission under contractH2020-787061 ANITA
\bibliography{multidomain}
\bibliographystyle{eamt22}
%\clearpage{}
%\appendix
%\input{appendix.tex}
\end{document}

>>>>

\author{Minh-Quang Pham\\
  Uni. Paris-Saclay,  CNRS, LISN \\
  F-91405 Orsay \\
  {\tt pham@limsi.fr}  \And
  \textbf{Josep Crego}\\
  SYSTRAN ,\\
  5 rue Feydeau, F-75002 Paris\\
  {\tt crego@systrangroup.com} \And
  Fran\c cois Yvon\\
  Uni. Paris-Saclay, CNRS, LISN\\
  F-91405 Orsay, France\\
  {\tt yvon@limsi.fr}}

@@

\section{Introduction}\label{sec:intro}
A typical setting in machine translation (MT) is to collect the largest possible collection of parallel data for the chosen language pair, with the intent to achieve optimal performance for the task of interest. In such situations, the training data distribution is opportunistic, while the test data distribution is chosen and fixed; a key aspect of training is then to mitigate the detrimental effects of a mismatch between these distributions. Single-source and multi-source\footnote{In this paper, multi-source DA means having multiple domains to adapt from; this setting differs from multi-source translation, where several \emph{source languages} are considered.} domain adaptation (DA) is a well-studied instance of this setting (see \cite{Chu2017comparison,Saunders21asurvey} for a review), and so is multi-domain (MD) learning \cite{Chu18multilingual,Zeng18multidomain,Jiang19multidomain,Pham21revisiting}. A related situation is multilingual MT \cite{Firat16multiway,Ha16towards,Johnson17google,Aharoni19massively},\fyDone{Fix ref}
% \revision{cite also \cite{Shen21source}} -> I think this is not so appropriate here
where the diversity of training data not only corresponds to variations in the topic, genre, or register but also in language.

@@

Several recent proposals
% \cite{Vanderwees17dynamic,Graves17automated,platanios19competence,Zhang19curriculum,Kumar19reinforcement,Wang20learning-multi,Wang20balancing,Wang20learning-multi}
explore ways to instead consider \emph{dynamic} data selection and sampling strategies: \newcite{Vanderwees17dynamic} and \newcite{Zhang19curriculum} construct a static curriculum, while \newcite{Wang20learning-multi} and \newcite{Wang20balancing} build curricula that automatically adapt to the training data. In this paper, we contribute to this line of research in several ways. 
\begin{itemize}
	\item First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short), a variant of Differentiable Data Selection (DDS) of \newcite{Wang20balancing}, initially applied to multilingual NMT, that simultaneously accounts for the domain adaptation and the multidomain adaptation problems.
	\item We show that MDAC achieves performance that compare to fine-tuning strategies for DA (\textsection~\ref{ssec:da}) and outperform some static data sampling strategies for multidomain settings (\ref{ssec:mda}).
	\item We show that our variant MDAC mitigates some failures of DDS in multidomain training.
	\item We illustrate the generality of differentiable data selection frameworks (both MDAC and DDS) on less common situations such as DA using unsupervised clustering (\textsection~\ref{ssec:clda}); DA using out-of-domain training data and small in-domain validation data (\textsection~\ref{ssec:uda}); and two-domain adaptation where the test distribution only mixes two of the training domain (\textsection~\ref{ssec:bida}).
\end{itemize}

@@

These situations have been amply documented from a theoretical perspective (eg.\ \cite{Mansour09multiple,Mansour09domainadaptation,Hoffman18algorithms}). A general recommendation in the DA setting is to adjust the sampling distribution used to optimize the system so as to compensate for the mismatch between $\mathcal{D}^s(x)$ and $\mathcal{D}^t(x)$. This can be approximated by reweighting instances, or more conveniently domains, which are selected during training with a probability $\lambda^{l}(d)$, with $\lambda^{l}(d) \neq \lambda^{s}(d)$.

@@

Multidomain NMT (MDMT) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDMT combine one or several ingredients: (a) the specialization of data representations \cite{Kobus17domaincontrol} or of sub-networks \cite{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \cite{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \cite{Jiang19multidomain}. Unsupervised MDMT is studied in \newcite{Farajian17multidomain}, as an instance of unsupervised DA.

Most approaches to adaptive/dynamic data selection take inspiration from \newcite{Bengio09curriculum}, where the notion of curriculum learning is introduced. CL relies on the notion of the ``easiness'' of a sample to schedule the presentation of training data so as to start with the easiest examples and end with the hardest. Various ways to automate CL in the framework of multi-armed bandits are explored in \cite{Graves17automated}, which has been an inspiration for our implementation. While the initial aim was primarily to improve and speed up training, CL has also proven useful for multidomain and multilingual MT, based on alternative definitions of ``easiness''.\fyDone{Ajouter Grave}
For instance, \newcite{Zhang19curriculum} study supervised DA and propose a curriculum approach which progressively augments the training data: early stages only use in-data, while less relevant\footnote{Domain distance is computed with Lewis-Moore scores (based on the cross-entropy of in-domain LM).} data are introduced in later stages. This is opposite to the policy of \newcite{Vanderwees17dynamic}, whose \emph{gradual fine-tuning} progressively increases the focus on in-domain data.\fyDone{These have not been compared? and also to what we do ?}

\newcite{Kumar19reinforcement} use reinforcement learning to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the devset loss that results from the current data selection strategy.\fyDone{Alert: what do we do during warm-up ?} This technique is applied to multilingual NMT in \cite{Kumar21learning}. \newcite{Zhou20uncertainty} propose another CL-based approach which relies on \emph{instance uncertainty} as a measure of their difficulty and presents data samples starting with the easiest. Another contribution of this work is a new stopping criterium. Closest to our problems, \newcite{Wang20learning-multi} adapt CL for multidomain NMT, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied to train a sequence of models. The one that maximizes the devset performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform multiple adaptation tasks for machine translation, ranging from supervised domain adaptation to multidomain NMT and unseen domain adaptation. In our experiments, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during the course of training, can be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.

@@

\bibliography{multidomain}
\bibliographystyle{eamt22}
% \clearpage{}

% \appendix
%\input{appendix.tex}
\end{document}
